Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-07-11 paper/analysis.tex
import changes from overleaf

%The idea behind the proof is to bound  to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality. We start by considering the residual of the 1-query Girard-Hutchinson estimator for a fixed matrix $\mtx{B}$: 
    \begin{equation}
        r(\mtx{B}, \vct{\psi}) := \Trace(\mtx{B}) - \vct{\psi}^{\top} \mtx{B} \vct{\psi}
    \end{equation}
%    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers
for a Gaussian random vector $\vct{\psi}$.
%
%    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. By \cite[Theorem 2.3]{boucheron-2013-concentration-inequalities} this implies for every $k \in \mathbb{N}$ that
    \begin{align*}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k}  \\
        &\le \big( k! 2^{4 k} + (2 k)! 2^{6 k}\big) \lVert \mtx{B} \rVert _F^{2 k} 
        \le \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{align*}
%     Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
%     \begin{equation}
%         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq 
%     \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} gives $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ of $r(\mtx{B}, \vct{\psi})$ is bounded by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    using the inequality $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, which can be verified using the monotonic decrease of the left-hand side with respect to $k \in \mathbb{N}$.

@@

The following result confirms the inituition that the Nystr\"om approximation is very well suited in the presence of decaying singular values: If the nuclear norm error to the best rank-$r$ approximation to $B(t)$ is bounded by $\varepsilon$ for every $t$ then the trace of the Nystr\"om approximation (with a bit of oversampling) results, with high probability, in an error that is only modestly larger than $\varepsilon$.

@@

\refthm{thm:hutchinson} has shown that the the Girard-Hutchinson estimator achieves a relative error $\varepsilon$ when using $\mathcal{O}(\varepsilon^{-2})$ queries. The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products with random vectors, \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$. This parallels existing results for the constant case
for Hutch++\cite[Theorem 3.1]{meyer-2021-hutch-optimal} and Nyström++\textcolor{red}{??? Include specific result ???}.

To do so, we mimic the proof of a similar bound for the Hutch++ estimator . Correspondingly, we will need to show that with $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2})$ random vectors, the Nyström approximation verifies $\int \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)\rVert _F~\mathrm{d}t \leq \varepsilon \int \Trace(\mtx{B}(t))~\mathrm{d}t$, regardless of the structure of the matrix $\mtx{B}(t)$. Just as in the corresponding proof for the Girard-Hutchinson estimator \refthm{thm:hutchinson}, we again bound the higher order moments of the approximation error for fixed $t$ and then use Minkowsi's integral inequality to carry over the results to the continuous case. We will do this in a sequence of three lemmas.

@@

%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}

\begin{proof}
\color{black}

    First, we temporarily omit the parameter dependence. Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B} \in \mathbb{R}^{n \times n}$, i.e. $\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$. We partition these matrices into
    \begin{equation}
        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
        \mtx{U} = \begin{bmatrix}
            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
        \end{bmatrix}
        \quad \text{and} \quad
        \mtx{\Lambda} =
        \begin{bmatrix}
            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
        \end{bmatrix}.
    \end{equation}
    Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are likewise standard Gaussian.

    We can apply \cite[Theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
    \begin{equation}
        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}
    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm.

    As in the proof of \cite[Lemma 3]{meyer-2021-hutch-optimal}, the first term -- and its moments -- are limited by
        \lVert \mtx{\Lambda}_2 \rVert _F
        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
        \leq \sqrt{ \frac{\Trace(\mtx{B})}{k} \Trace(\mtx{B})}
        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{B}).
        \label{equ:nystrom-proof-frobenius-trace}

>>>>

%The idea behind the proof is to bound  to extend the results to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a diagonal embedding trick to carry the result over to the general case.
\begin{proof} 
\textcolor{red}{??? I feel that the block diagonal embedding trick should come first, like in the proof of Meyer et al. Hopefully, this gets rid of the additional $\log \delta$ factor ???}
The proof proceeds by bounding moments for fixed values of $t$ and then use Minkowski's integral inequality. We start by considering the residual of the 1-query Girard-Hutchinson estimator for a fixed matrix $\mtx{B}$: 
    \begin{equation}
        r(\mtx{B}, \vct{\psi}) := \Trace(\mtx{B}) - \vct{\psi}^{\top} \mtx{B} \vct{\psi}
    \end{equation}
%    for a symmetric parameter-dependent matrix $\mtx{B}(t)$ over the real numbers
for a Gaussian random vector $\vct{\psi}$.
%
%    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence.
The proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} shows that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. By \cite[Theorem 2.3]{boucheron-2013-concentration-inequalities} this implies for every $k \in \mathbb{N}$ that
    \begin{align*}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k}  \\
        &\le \big( k! 2^{4 k} + (2 k)! 2^{6 k}\big) \lVert \mtx{B} \rVert _F^{2 k} 
        \le \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{align*}
%     Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
%     \begin{equation}
%         \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq 
%     \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} gives $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. \textcolor{red}{??? this norm should be defined more prominently somewhere else ???} Consequently, the $L^{2k}$ norm $\mathbb{E}^{2k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{2k} \right] \right)^{\sfrac{1}{2k}}$ of $r(\mtx{B}, \vct{\psi})$ is bounded by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    using the inequality $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, which can be verified using the monotonic decrease of the left-hand side with respect to $k \in \mathbb{N}$.

@@

A consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om approximation is very well suited in the presence of decaying singular values: If the nuclear norm error to the best rank-$r$ approximation to $B(t)$ is bounded by $\varepsilon$ for every $t$ then the trace of the Nystr\"om approximation (with a bit of oversampling) results, with high probability, in an error that is only modestly larger than $\varepsilon$.

@@

\refthm{thm:hutchinson} shows that the the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries. The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$. This parallels existing results for the constant case
for Hutch++\cite[Theorem 3.1]{meyer-2021-hutch-optimal} and Nyström++\textcolor{red}{??? Include specific result ???}. Indeed, our proof follows the strategy from~\cite{meyer-2021-hutch-optimal}, by first showing that the parameter-dependent Nyström approximation achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries.

@@

%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
\begin{proof}
For fixed $t$, consider the spectral decomposition $\mtx{B}(t) = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ with  
$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\lambda_1 \ge \dots \ge \lambda_n \ge 0$. Partitioning
    \begin{equation*}
        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
        \mtx{U} = \begin{bmatrix}
            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
        \end{bmatrix}
        \quad \text{and} \quad
        \mtx{\Lambda} =
        \begin{bmatrix}
            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
        \end{bmatrix},
    \end{equation*}
    we set
    $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$ yields the bound
    \begin{equation}
        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}(t)} \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}
    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm. Standard arguments (see, e.g., the proof of~\cite[Lemma 3]{meyer-2021-hutch-optimal}) provide a simple bound on the first term:
    \begin{equation}
        \lVert \mtx{\Lambda}_2 \rVert _F
        \leq \sqrt{ \lambda_{k+1} (  \lambda_{k+1} + \cdots + \lambda_{n})}
        \leq \sqrt{ \Trace(\mtx{B}) / k \cdot \Trace(\mtx{B})}
        \leq \Trace(\mtx{B}) / \sqrt{k}.
        \label{equ:nystrom-proof-frobenius-trace}
    \end{equation}
    \textcolor{red}{??? STOPPED HERE: Not clear to me why $n_\Omega/4$ should be the optimal exponent for the tail bound ???}

### 2024-07-11 paper/paper.tex
import changes from overleaf

\textcolor{red}{TODO for Fabio:
\begin{itemize}
 \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style. The paper needs to get strictly below 30 pages prior to submission. \textcolor{green}{Journal abbreviations, max 2 authors, omit superfluous fields $\to$ 3 pages.}
 Journal abbreviations and removing superfluous fields is good. max 2 authors is not coming in math, please reverse this. \textcolor{green}{Done.} The text style is a bit too generous but no need to change this now. \textcolor{green}{Ok! Let's talk about it at some point.}
\end{itemize}
}
\color{blue}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
%\input{numresults.tex}

@@

\appendix

\section{Proofs}

\textcolor{red}{??? Need to check carefully that this result is not known ???} \textcolor{green}{I haven't found anything so far. The fact that also Bujanović/Grubišić/Kressner/Lam in Lemma 20 use a very similar technique, but merely to bound $\Gamma(\frac{p+1}{2})$ makes me suspect there's not much, except for Stirling approximation at the cost of some simplicity. Will keep investigating.}

\begin{lemma}{Moment bound of $\chi^2$-random variable}{gamma}
    Let $X$ follow a $\chi^2$ distribution with $k \in \mathbb{N}$ degrees of freedom. Then for all $p \in \mathbb{N}$ it holds
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}[X] \leq k + p
        %\frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \leq \left( \frac{k + p}{2} \right)^{\sfrac{p}{2}}.
    \end{equation}
\end{lemma}

\begin{proof}
    From \cite[Theorem 3.3.2]{hogg-2013-introduction-mathematical} we know
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}[X] = 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}}.
    \end{equation}
    We first treat the case $k=p=1$, for which indeed
    \begin{equation}
        2 \left( \frac{\Gamma(\frac{1+1}{2})}{\Gamma(\frac{1}{2})} \right)^2 = \frac{2}{\pi} \leq 2 = 1 + 1
    \end{equation}
    So we now assume $k + p \geq 3$. We can use a telescoping product and the property $\Gamma(z+1)/\Gamma(z) = z$ to bound
    \begin{align}
        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
        &= \frac{\Gamma(\frac{k}{2} + 1)}{\Gamma(\frac{k}{2})} \frac{\Gamma(\frac{k}{2} + 2)}{\Gamma(\frac{k}{2} + 1)} \cdots \frac{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor - 1)} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
        &= \left(\frac{k}{2}\right) \left(\frac{k}{2} + 1\right) \cdots \left( \frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1 \right) \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)} \notag \\
        &\leq \left(\frac{k}{2} + \bigg\lfloor \frac{p}{2} \bigg\rfloor - 1\right)^{\lfloor \sfrac{p}{2} \rfloor} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k}{2} + \lfloor \frac{p}{2} \rfloor)}
    \end{align}
    If $p$ is even, then $\lfloor p/2 \rfloor = p/2$ and hence
    \begin{equation}
        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} 
        \leq \left( \frac{k + p}{2} - 1 \right)^{\sfrac{p}{2}}
        \leq \left( \frac{k + p}{2}\right)^{\sfrac{p}{2}}.
    \end{equation}
    If $p$ is odd, then $\lfloor p/2 \rfloor = (p - 1)/2$, from which follows with Gautschi's inequality \cite{kershaw-1983-extensions-gautschi}
    \begin{equation}
        \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})}
        \leq \left(\frac{k + p}{2} - \frac{3}{2} \right)^{ \frac{p - 1}{2}} \frac{\Gamma( \frac{k + p}{2})}{\Gamma(\frac{k + p}{2} - \frac{1}{2})}
        \leq \left(\frac{k + p}{2}\right)^{ \frac{p - 1}{2}} \left(\frac{k + p}{2}\right)^{\sfrac{1}{2}}
        = \left(\frac{k + p}{2}\right)^{ \sfrac{p}{2}}
    \end{equation}
    Hence,
    \begin{equation}
        \mathbb{E}^{\sfrac{p}{2}}[X] \leq 2 \left( \frac{k + p}{2} \right) \leq k + p.
    \end{equation}
\end{proof}

>>>>

\textcolor{red}{TODO for Fabio:
\begin{itemize}
 \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style. The paper needs to get strictly below 30 pages prior to submission. \textcolor{green}{Journal abbreviations, max 2 authors, omit superfluous fields $\to$ 3 pages.}
 Journal abbreviations and removing superfluous fields is good. max 2 authors is not coming in math, please reverse this. \textcolor{green}{Done.} The text style is a bit too generous but no need to change this now. \textcolor{green}{Ok! Let's talk about it at some point.}
\end{itemize}
}
\color{blue}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
\input{numresults.tex}

@@

\input{appendix.tex}
